Mapping mutational effects along the evolutionary landscape of HIV envelope

### 2018-02-15 paper/manuscript.tex
added blank line to manuscript

\begin{table}
\begin{fullwidth}
{\centering \include{figures/gamma_omega_modelcomparison}}
\caption{\label{tab:phydms}
Evolutionary models informed by the deep mutational scanning describe HIV's evolution in nature much better than a standard substitution model.
Shown are the results of maximum likelihood fitting of substitution models to the clade A phylogeny in Figure~\ref{fig:tree}.
Experimentally informed codon models~\citep[ExpCM,][]{hilton2017phydms} utilizing the across-replicate average of the deep mutational scanning describe Env's natural evolution far better than a standard codon substitution model~\citep[the M5 model of][]{yang2000codon} as judged by comparing the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}.
Both ExpCM models have a stringency parameter $>$1.
All models draw $\omega$ from a gamma distribution, and the table shows the mean ($\overline{\omega}$) and shape parameters ($\omega_{\alpha}$ and $\omega_{\beta}$) of this distribution.
The last two columns show the number of sites evolving faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected at a false discovery rate of 0.05, as determined using the approach in \citet{bloom2017identification}.
Analyses were performed using \texttt{phydms} (\url{http://jbloomlab.github.io/phydms/}).
Table~\ref{tab:phydms}-source-data~\ref{tabledata:phydms} shows the results for additional substitution models.
}
\tabledata{\label{tabledata:phydms}Results for phylogenetic models where $\omega$ is not drawn from a gamma-distribution or where the preferences are averaged across sites to eliminate the site specificity are in \texttt{modelcomparison.md}.}
\end{fullwidth}
\end{table}

Table~\ref{tab:phydms} shows that phylogenetic models informed by the deep mutational scanning of either BG505 or BF520 describe the natural evolution of Env vastly better than a standard codon substitution model.
In addition to simply comparing model fits, we can interpret the model parameters.
ExpCM's have a stringency parameter that relates selection in the experiments to that in nature.
A stringency parameter $>$1 indicates that natural selection prefers the same amino acids as the experiments, but with greater stringency~\citep{hilton2017phydms}.
Both ExpCM's have stringency parameters $>$1 (Table~\ref{tab:phydms}) -- a finding that makes sense, since the stop-codon analysis in the previous section suggests that the experimental selections are more lax than natural selection on HIV.
We can also interpret the $\omega$ parameter.
For standard codon substitution models, $\omega$ is the rate of fixation of nonsynonymous mutations relative to synonymous ones -- and the gene-wide average $\omega$ is almost always $<$1, since purifying selection purges many functionally deleterious amino-acid mutations even for adaptively evolving proteins~\citep{murrell2015gene}.
Indeed, Table~\ref{tab:phydms} shows that Env's gene-wide average $\omega$ is $<$1 for a standard model.
But for ExpCM's, $\omega$ is the relative rate of nonsynonymous to synonymous substitutions \emph{after} accounting for functional constraints measured in the deep mutational scanning~\citep{bloom2017identification}.
Table~\ref{tab:phydms} shows that the ExpCM's have a gene-wide average $\omega$ that is $>$1, indicating that external selection (e.g., from immunity) drives Env to fix amino-acid mutations faster than expected under a null model that only accounts for functional constraints.

\begin{figure}
\begin{fullwidth}
{\includegraphics[width=1.3\textwidth]{figures/BG505_prefs.pdf}}
\caption{\label{fig:BG505prefs}
Amino-acid preferences for the BG505 Env.
At each site, the height of the letter is proportional for that site's preference for that amino acid.
The top color bar indicates the region of Env (gp120 variable loop, gp120 not variable loop, or gp41).
The lower color bar indicates the evidence that the site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experiments~\citep[see][]{bloom2017identification}.
The letters above the logos indicate the wildtype amino acid in BG505.
Sites are numbered using the HXB2 scheme~\citep{korber1998numbering}).
This logo plot shows the site-specific amino-acid preferences for BG505 after averaging the replicates and re-scaling by the stringency parameter in Table \ref{tab:phydms}.
The figure was generated using \texttt{dms\_tools2}~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}, which in turn utilizes \texttt{weblogo}~\citep{crooks2004weblogo}.
}
\figdata{The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BG505\_prefs.csv.}}
\figdata{The sequence of BG505 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BG505\_to\_HXB2.csv}.}
\figdata{The $\omega_r$ values and associated $P$-values for BG505 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
\end{fullwidth}
\end{figure}

\begin{figure}
\begin{fullwidth}
{\includegraphics[width=1.3\textwidth]{figures/BF520_prefs.pdf}}
\caption{\label{fig:BF520prefs}
Amino-acid preferences for the BF520 Env.
This figure is the same as Figure~\ref{fig:BG505prefs} except that it shows the data for BF520 instead of BG505.
}
\figdata{The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BF520\_prefs.csv.}}
\figdata{The sequence of BF520 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BF520\_to\_HXB2.csv}.}
\figdata{The $\omega_r$ values and associated $P$-values for BF520 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
\end{fullwidth}
\end{figure}

A more qualitative way to assess if the deep mutational scanning authentically describes selection on Env function is to visually compare the measurements with existing knowledge.
Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} show the across-replicate average of the amino-acid preferences for each Env after re-scaling by the stringency parameters in Table~\ref{tab:phydms}.
(Note that throughout the rest of this paper, we use preferences re-scaled by these stringency parameters.)
At sites of known functional importance, these preferences are usually consistent with prior knowledge.
For instance, residues T257, D368, E370, W427, and D457 are important for Env binding to CD4~\citep{olshevsky1990identification}, and all these amino acids are highly preferred in our deep mutational scanning (Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}).
Likewise, Env has 10 disulfide bonds (linking sites 54-74, 119-205, 126-196, 131-157, 218-247, 228-239, 296-331, 378-445, 385-418, and 598-604), most of which are important for function~\citep{van2008only} -- and the cysteines at these sites are highly preferred in our deep mutational scanning.
The deep mutational scanning is also consistent with prior knowledge about sites that are tolerant of mutations.
For instance, Env has five variable loops that mostly evolve under weak constraint in nature~\citep{starcich1986identification,zolla2010structure} -- and most sites in these loops are mutationally tolerant in our deep mutational scanning (see sites indicated by gray overlay bars in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}, such as 132 to 195).
It is beyond the scope of this paper to catalog associations between our measurements and all other prior mutational studies of Env, but the concordance of our findings with the above mutational studies, and the fact that our data improve phylogenetic models of Env's natural evolution, suggest that our measurements authentically describe functional selection on Env.

\subsection{Shifts in amino-acid preferences between BG505 and BF520}
The most fundamental question that we seek to address is how similar the amino-acid preferences are between the two Envs.
We have already noted that Figure~\ref{fig:mutfreqs}B shows that the preferences are more correlated for replicate measurements on the same Env than for replicate measurements on different Envs.
However, simply comparing correlation coefficients does not identify specific sites where mutational effects have shifted, nor does it quantify the magnitude of any shifts.

@@

We therefore used a more rigorous approach to identify sites where the amino-acid preferences differ between BG505 and BF520 by an amount that exceeds the noise in our experiments.
We first re-scaled the preferences from each experimental replicate by the stringency parameter for that Env from Table~\ref{tab:phydms} to calibrate all measurements to the stringency of natural selection.
We then identified the 659 sites in the mutagenized regions of Env that are pairwise alignable between BG505 and BF520 (Figure~\ref{fig:prefsdist}-source~data~\ref{figdata:prefsdist}).
For each site, we calculated the shift in amino-acid preferences between Envs using an approach similar to that of \citet{doud2015site} as illustrated in Figure~\ref{fig:prefsdist}A.
This approach calculates the magnitude of the shift after correcting for experimental noise by comparing the differences in preferences between replicates for BG505 and BF520 to the differences between replicates for the same Env.
Figure~\ref{fig:prefsdist}A shows this calculation for a site that has not shifted (site 598, which strongly prefers cysteine in both Envs), the most shifted site (512, which shifts from being mutationally tolerant in BG505 to strongly preferring alanine in BF520), and two other sites with more intermediate behaviors.

The overall distribution of shifts between BG505 and BF520 is shown in Figure~\ref{fig:prefsdist}B.
Most sites have relatively small shifts (close to zero), although there is a long tail of sites with large shifts.
This tail reaches its upper value with site 512, which has a shift of 0.52 out of a maximum possible of 1.0.
How should we interpret this distribution -- have mutational effects shifted a lot, or not very much?
We can establish an upper-bound for how much sites might shift by comparing Env to a \emph{non}-homologous protein.
Figure~\ref{fig:prefsdist}B shows the distribution of shifts when comparing Env to influenza's hemagglutinin protein, which has previously had its amino-acid preferences measured by deep mutational scanning~\citep{doud2016accurate} .
Most sites have large shifts between Env and hemagglutinin, with the typical shift being $\sim$0.4 and some approaching the maximum value of 1.0.
We can also establish a lower-bound by creating a null distribution for the expected shifts if all differences are simply due to experimental noise.
This null distribution is created by randomizing the experimental replicates among Envs.
Figure~\ref{fig:prefsdist}B shows that the null distribution is more peaked at zero than the real distribution, and does not have the same prominent tail of sites with large shifts.
The answer to the question of how much mutational effects have shifted is therefore nuanced: they have substantially shifted at some sites, but remain vastly more similar between the two Envs than between two unrelated proteins.

We can use the null distribution to identify sites where the shifts between BG505 and BF520 are significantly larger than the noise in our experiments (Figure~\ref{fig:prefsdist}B).
There are 30 such sites at a false discovery rate of 0.1.
Figure~\ref{fig:prefsdist}C shows the amino-acid preferences of these significantly shifted sites for each Env.
For the majority of shifted sites, one Env prefers a specific amino acid whereas the other Env tolerates many amino acids; for instance, see sites 512, 516, 599, 165, 605 and 505 in Figure~\ref{fig:prefsdist}C.
Such broadening and narrowing of a site's mutational tolerance is frequently linked to changes in protein stability, with a more stable protein typically being more mutationally tolerant~\citep{wang2002evolution,bloom2006protein,gong2013stability,kumar2017stability}.
Work with engineered Env protein in the form of ``SOSIP'' trimer~\citep{binley2000recombinant,sanders2002stabilization} has shown that BG505 SOSIP is more thermostable than BF520 SOSIP~\citep{verkerke2016epitope}.
Consistent with this fact, sites with altered mutational tolerance are often (although not always, see sites 165 and 520 in Figure~\ref{fig:prefsdist}C) more mutationally tolerant in BG505.

However, not all of the significantly shifted sites show a simple pattern of broadening or narrowing of mutational tolerance.
For instance, site 288 does not alter its mutational tolerance but rather flips its rather narrow amino-acid preference from phenylalanine in BG505 to leucine in BF520 (Figure~\ref{fig:prefsdist}C).
Thus, there is variation in both the extent and types of shifts observed.

@@

\begin{figure}
{\bf \Large A} \\
\includegraphics[width=0.8\textwidth]{figures/shifts_on_structure/shifts_on_structure.pdf}
\\
{\bf \Large B} \hspace{0.28\textwidth} {\bf \Large C} \hspace{0.28\textwidth} {\bf \Large D} \\
\includegraphics[width=0.28\textwidth]{figures/rsa_vs_shifts.pdf}
\hspace{0.01\textwidth}
\includegraphics[width=0.28\textwidth]{figures/shifts_proximity.pdf}
\hspace{0.01\textwidth}
\includegraphics[width=0.4\textwidth]{figures/conservation_vs_shifts.pdf} 
\caption{\label{fig:shiftfeatures}
Characteristics of significantly shifted sites.
{\bf (A)} One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from gray to orange according to the magnitude of the mutational shift at each site (orange indicates large shift).
Sites that are significantly shifted according to Figure~\ref{fig:prefsdist}B are in spheres, and all other sites are in cartoon representation.
{\bf (B)} 
There is no significant difference in the relative solvent accessibility of sites that have and have not undergone significant shifts.
The absolute solvent accessibility of each site was calculated using DSSP~\citep{kabsch1983dictionary} and normalized to a relative solvent accessibility using the absolute accessibilities from \citet{tien2013maximum}.
{\bf (C)} 
Sites of significant shifts are clustered in Env's structure.
Shown is the distance of each significantly shifted and not-shifted site to the closest other shifted site.
{\bf (D)}
Large mutational shifts are \emph{not} strongly enriched at sites that have substituted between BG505 and BF520, or at sites that contact sites that have substituted. 
The plot shows the magnitudes of the shifts among the 83 crystallographically resolved sites that have substituted between BG505 and BF520, the 234 non-substituted sites that physically contact a substitution in the Env structure (any non-hydrogen atom within 3.5\angstrom), and all other sites.
Figure~\ref{fig:shiftfeatures}-Figure~supplement~\ref{figsupp:subs_shifts} shows that there is a modest borderline-significant tendency of significantly shifted sites to have substituted.
All panels only show the 577 sites that are resolved in the crystal structure.
Of the 30 significantly shifted sites in Figure~\ref{fig:prefsdist}B, only 28 are crystallographically resolved.
Structural distances and solvent accessibilities were calculated using all monomers in the trimer.
$P$ values were calculated using the Mann-Whitney $U$ test.
}
\figsupp[\label{figsupp:subs_shifts}Statistical testing of whether significantly shifted sites are more likely to have substituted between BG505 and BF520.]
{The sites of significant shifts as determined in Figure~\ref{fig:prefsdist}B are somewhat more likely to be among the sites that have substituted between BG505 and BF520.
However, this association is only borderline statistically significant, with $P = 0.055$ using a Fisher's exact test.}
{\input{figures/shifts_vs_subs_table}}
\end{figure}

What distinguishes the sites that have undergone significant shifts?
Figure~\ref{fig:shiftfeatures}A shows the locations of the shifted sites on the crystal structure of Env.
There is no visually obvious tendency for shifted sites to preferentially be on Env's surface or in its core, and a statistical analysis (Figure~\ref{fig:shiftfeatures}B) finds no association between a site's relative solvent accessibility and whether its amino-acid preferences have shifted.
However, Figure~\ref{fig:shiftfeatures}A does suggest that the sites of significant shifts tend to cluster in Env's structure, and a statistical analysis confirms that this is the case (Figure~\ref{fig:shiftfeatures}C).
Therefore, the factors that drive shifts in Env's mutational tolerance often affect physically interacting clusters of residues in a coordinated fashion.

An obvious hypothesis is that strongly shifted sites have substituted between BG505 and BF520, or physically contact such substitutions.
According to this hypothesis, substitutions would alter the local physicochemical environment of the substituted site and its neighbors, thereby shifting the amino-acid preferences of sites in the physical cluster.
But surprisingly, the typical magnitude of shifts is not significantly larger at sites that have substituted, or at sites that contact sites that have experienced substitutions (Figure~\ref{fig:shiftfeatures}C).
There is a borderline trend for the significantly shifted sites to be more likely to have substituted between BG505 and BF520 (Figure~\ref{fig:shiftfeatures}-Figure~supplement~\ref{figsupp:subs_shifts}), but most shifted sites have not substituted (only 8 of the 30 shifted sites differ in amino-acid identity between the two Envs).
The lack of strong enrichment in shifts at substituted sites contrasts with previous protein-wide experimental~\citep{doud2015site} and simulation-based~\citep{pollock2012amino,shah2015contingency} studies of shifting amino-acid preferences, which found that shifts were dramatically more pronounced at sites of substitutions.
The difference may arise because these earlier studies examined proteins that are fairly conformationally static (absolutely so in the case of the simulations).
On the other hand, Env is extremely complex and conformationally dynamic~\citep{munro2014conformational,ozorowski2017open}, which may increase the opportunities for long-range epistasis to enable substitutions at one site to shift the amino-acid preferences of distant sites.

@@

Figure~\ref{fig:entrenchment} shows the distribution of effects for mutating all sites that differ between BG505 and BF520 to the identity in the other Env.
As expected under entrenchment, the average effect of these mutations is deleterious -- although there are a substantial number of sites where the mutational flips are not deleterious. 
We can get some sense of the magnitude of the entrenchment by comparing the effects of the BG505$\leftrightarrow$BF520 mutations to the distribution of effects of all possible amino-acid mutations (Figure~\ref{fig:entrenchment}).
This comparison shows that even unfavorable inter-Env mutational flips are generally more favorable than random amino-acid mutations.
Therefore, entrenchment occurs for some but not all substitutions that distinguish BG505 and BF520, and the magnitude of entrenchment is less than the effect of a typical random mutation.
Entrenchment of substitutions therefore contributes to some of the mutational shifts.
But given that many of these shifts occur at sites that do not even differ between the Envs (Figure~\ref{fig:shiftfeatures}D), entrenchment of substitutions is clearly not the only cause of the shifting amino-acid preferences.

@@

Our results show that amino-acid preference maps of Env also have a useful level of conservation for many purposes.
From a qualitative perspective, the amino-acid preferences look mostly similar between BG505 and BF520, and so provide a valuable reference for estimating which mutations are likely to be tolerated at each site in diverse HIV strains.
Indeed, we anticipate that the complete maps of mutational effects in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} will be useful for future sequence-structure-function studies.
From an analytical perspective, a powerful use of our maps is to identify sites that evolve differently in nature than is required by the simple selection for viral growth imposed in our experiments -- and the identified sites are largely the same regardless of whether the analysis uses an amino-acid preference map from BG505 or BF520.

>>>>

\begin{table}
\begin{fullwidth}
{\centering \include{figures/gamma_omega_modelcomparison}}
\caption{\label{tab:phydms}
Evolutionary models informed by the deep mutational scanning describe HIV's evolution in nature much better than a standard substitution model.
Shown are the results of maximum likelihood fitting of substitution models to the clade A phylogeny in \FIG{tree}.
Experimentally informed codon models~\citep[ExpCM,][]{hilton2017phydms} utilizing the across-replicate average of the deep mutational scanning describe Env's natural evolution far better than a standard codon substitution model~\citep[the M5 model of][]{yang2000codon} as judged by comparing the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}.
Both ExpCM models have a stringency parameter $>$1.
All models draw $\omega$ from a gamma distribution, and the table shows the mean ($\overline{\omega}$) and shape parameters ($\omega_{\alpha}$ and $\omega_{\beta}$) of this distribution.
The last two columns show the number of sites evolving faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected at a false discovery rate of 0.05, as determined using the approach in \citet{bloom2017identification}.
Analyses were performed using \texttt{phydms} (\url{http://jbloomlab.github.io/phydms/}).
\TABLE{phydms}-source-data~\ref{tabledata:phydms} shows the results for additional substitution models.
}
\tabledata{\label{tabledata:phydms}Results for phylogenetic models where $\omega$ is not drawn from a gamma-distribution or where the preferences are averaged across sites to eliminate the site specificity are in \texttt{modelcomparison.md}.}
\end{fullwidth}
\end{table}

\TABLE{phydms} shows that phylogenetic models informed by the deep mutational scanning of either BG505 or BF520 describe the natural evolution of Env vastly better than a standard codon substitution model.
In addition to simply comparing model fits, we can interpret the model parameters.
ExpCM's have a stringency parameter that relates selection in the experiments to that in nature.
A stringency parameter $>$1 indicates that natural selection prefers the same amino acids as the experiments, but with greater stringency~\citep{hilton2017phydms}.
Both ExpCM's have stringency parameters $>$1 (\TABLE{phydms}) -- a finding that makes sense, since the stop-codon analysis in the previous section suggests that the experimental selections are more lax than natural selection on HIV.
We can also interpret the $\omega$ parameter.
For standard codon substitution models, $\omega$ is the rate of fixation of nonsynonymous mutations relative to synonymous ones -- and the gene-wide average $\omega$ is almost always $<$1, since purifying selection purges many functionally deleterious amino-acid mutations even for adaptively evolving proteins~\citep{murrell2015gene}.
Indeed, \TABLE{phydms} shows that Env's gene-wide average $\omega$ is $<$1 for a standard model.
But for ExpCM's, $\omega$ is the relative rate of nonsynonymous to synonymous substitutions \emph{after} accounting for functional constraints measured in the deep mutational scanning~\citep{bloom2017identification}.
\TABLE{phydms} shows that the ExpCM's have a gene-wide average $\omega$ that is $>$1, indicating that external selection (e.g., from immunity) drives Env to fix amino-acid mutations faster than expected under a null model that only accounts for functional constraints.

\begin{figure}
\begin{fullwidth}
{\includegraphics[width=1.3\textwidth]{figures/BG505_prefs.pdf}}
\caption{\label{fig:BG505prefs}
Amino-acid preferences for the BG505 Env.
At each site, the height of the letter is proportional for that site's preference for that amino acid.
The top color bar indicates the region of Env (gp120 variable loop, gp120 not variable loop, or gp41).
The lower color bar indicates the evidence that the site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experiments~\citep[see][]{bloom2017identification}.
The letters above the logos indicate the wildtype amino acid in BG505.
Sites are numbered using the HXB2 scheme~\citep{korber1998numbering}).
This logo plot shows the site-specific amino-acid preferences for BG505 after averaging the replicates and re-scaling by the stringency parameter in \TABLE{phydms}.
The figure was generated using \texttt{dms\_tools2}~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}, which in turn utilizes \texttt{weblogo}~\citep{crooks2004weblogo}.
}
\figdata{The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BG505\_prefs.csv.}}
\figdata{The sequence of BG505 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BG505\_to\_HXB2.csv}.}
\figdata{The $\omega_r$ values and associated $P$-values for BG505 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
\end{fullwidth}
\end{figure}

\begin{figure}
\begin{fullwidth}
{\includegraphics[width=1.3\textwidth]{figures/BF520_prefs.pdf}}
\caption{\label{fig:BF520prefs}
Amino-acid preferences for the BF520 Env.
This figure is the same as \FIG{BG505prefs} except that it shows the data for BF520 instead of BG505.
}
\figdata{The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BF520\_prefs.csv.}}
\figdata{The sequence of BF520 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BF520\_to\_HXB2.csv}.}
\figdata{The $\omega_r$ values and associated $P$-values for BF520 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
\end{fullwidth}
\end{figure}

A more qualitative way to assess if the deep mutational scanning authentically describes selection on Env function is to visually compare the measurements with existing knowledge.
\FIG{BG505prefs} and \FIG{BF520prefs} show the across-replicate average of the amino-acid preferences for each Env after re-scaling by the stringency parameters in \TABLE{phydms}.
(Note that throughout the rest of this paper, we use preferences re-scaled by these stringency parameters.)
At sites of known functional importance, these preferences are usually consistent with prior knowledge.
For instance, residues T257, D368, E370, W427, and D457 are important for Env binding to CD4~\citep{olshevsky1990identification}, and all these amino acids are highly preferred in our deep mutational scanning (\FIG{BG505prefs} and \FIG{BF520prefs}).
Likewise, Env has 10 disulfide bonds (linking sites 54-74, 119-205, 126-196, 131-157, 218-247, 228-239, 296-331, 378-445, 385-418, and 598-604), most of which are important for function~\citep{van2008only} -- and the cysteines at these sites are highly preferred in our deep mutational scanning.
The deep mutational scanning is also consistent with prior knowledge about sites that are tolerant of mutations.
For instance, Env has five variable loops that mostly evolve under weak constraint in nature~\citep{starcich1986identification,zolla2010structure} -- and most sites in these loops are mutationally tolerant in our deep mutational scanning (see sites indicated by gray overlay bars in \FIG{BG505prefs} and \FIG{BF520prefs}, such as 132 to 195).
It is beyond the scope of this paper to catalog associations between our measurements and all other prior mutational studies of Env, but the concordance of our findings with the above mutational studies, and the fact that our data improve phylogenetic models of Env's natural evolution, suggest that our measurements authentically describe functional selection on Env.

\subsection{Shifts in amino-acid preferences between BG505 and BF520}
The most fundamental question that we seek to address is how similar the amino-acid preferences are between the two Envs.
We have already noted that \FIG{mutfreqs}B shows that the preferences are more correlated for replicate measurements on the same Env than for replicate measurements on different Envs.
However, simply comparing correlation coefficients does not identify specific sites where mutational effects have shifted, nor does it quantify the magnitude of any shifts.

@@

We therefore used a more rigorous approach to identify sites where the amino-acid preferences differ between BG505 and BF520 by an amount that exceeds the noise in our experiments.
We first re-scaled the preferences from each experimental replicate by the stringency parameter for that Env from \TABLE{phydms} to calibrate all measurements to the stringency of natural selection.
We then identified the 659 sites in the mutagenized regions of Env that are pairwise alignable between BG505 and BF520 (\FIG{prefsdist}-source~data~\ref{figdata:prefsdist}).
For each site, we calculated the shift in amino-acid preferences between Envs using an approach similar to that of \citet{doud2015site} as illustrated in \FIG{prefsdist}A.
This approach calculates the magnitude of the shift after correcting for experimental noise by comparing the differences in preferences between replicates for BG505 and BF520 to the differences between replicates for the same Env.
\FIG{prefsdist}A shows this calculation for a site that has not shifted (site 598, which strongly prefers cysteine in both Envs), the most shifted site (512, which shifts from being mutationally tolerant in BG505 to strongly preferring alanine in BF520), and two other sites with more intermediate behaviors.

The overall distribution of shifts between BG505 and BF520 is shown in \FIG{prefsdist}B.
Most sites have relatively small shifts (close to zero), although there is a long tail of sites with large shifts.
This tail reaches its upper value with site 512, which has a shift of 0.52 out of a maximum possible of 1.0.
How should we interpret this distribution -- have mutational effects shifted a lot, or not very much?
We can establish an upper-bound for how much sites might shift by comparing Env to a \emph{non}-homologous protein.
\FIG{prefsdist}B shows the distribution of shifts when comparing Env to influenza's hemagglutinin protein, which has previously had its amino-acid preferences measured by deep mutational scanning~\citep{doud2016accurate} .
Most sites have large shifts between Env and hemagglutinin, with the typical shift being $\sim$0.4 and some approaching the maximum value of 1.0.
We can also establish a lower-bound by creating a null distribution for the expected shifts if all differences are simply due to experimental noise.
This null distribution is created by randomizing the experimental replicates among Envs.
\FIG{prefsdist}B shows that the null distribution is more peaked at zero than the real distribution, and does not have the same prominent tail of sites with large shifts.
The answer to the question of how much mutational effects have shifted is therefore nuanced: they have substantially shifted at some sites, but remain vastly more similar between the two Envs than between two unrelated proteins.

We can use the null distribution to identify sites where the shifts between BG505 and BF520 are significantly larger than the noise in our experiments (\FIG{prefsdist}B).
There are 30 such sites at a false discovery rate of 0.1.
\FIG{prefsdist}C shows the amino-acid preferences of these significantly shifted sites for each Env.
For the majority of shifted sites, one Env prefers a specific amino acid whereas the other Env tolerates many amino acids; for instance, see sites 512, 516, 599, 165, 605 and 505 in \FIG{prefsdist}C.
Such broadening and narrowing of a site's mutational tolerance is frequently linked to changes in protein stability, with a more stable protein typically being more mutationally tolerant~\citep{wang2002evolution,bloom2006protein,gong2013stability,kumar2017stability}.
Work with engineered Env protein in the form of ``SOSIP'' trimer~\citep{binley2000recombinant,sanders2002stabilization} has shown that BG505 SOSIP is more thermostable than BF520 SOSIP~\citep{verkerke2016epitope}.
Consistent with this fact, sites with altered mutational tolerance are often (although not always, see sites 165 and 520 in \FIG{prefsdist}C) more mutationally tolerant in BG505.

However, not all of the significantly shifted sites show a simple pattern of broadening or narrowing of mutational tolerance.
For instance, site 288 does not alter its mutational tolerance but rather flips its rather narrow amino-acid preference from phenylalanine in BG505 to leucine in BF520 (\FIG{prefsdist}C).
Thus, there is variation in both the extent and types of shifts observed.

@@

\begin{figure}
{\bf \Large A} \\
\includegraphics[width=0.8\textwidth]{figures/shifts_on_structure/shifts_on_structure.pdf}
\\
{\bf \Large B} \hspace{0.28\textwidth} {\bf \Large C} \hspace{0.28\textwidth} {\bf \Large D} \\
\includegraphics[width=0.28\textwidth]{figures/rsa_vs_shifts.pdf}
\hspace{0.01\textwidth}
\includegraphics[width=0.28\textwidth]{figures/shifts_proximity.pdf}
\hspace{0.01\textwidth}
\includegraphics[width=0.4\textwidth]{figures/conservation_vs_shifts.pdf} 
\caption{\label{fig:shiftfeatures}
Characteristics of significantly shifted sites.
{\bf (A)} One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from gray to orange according to the magnitude of the mutational shift at each site (orange indicates large shift).
Sites that are significantly shifted according to \FIG{prefsdist}B are in spheres, and all other sites are in cartoon representation.
{\bf (B)} 
There is no significant difference in the relative solvent accessibility of sites that have and have not undergone significant shifts.
The absolute solvent accessibility of each site was calculated using DSSP~\citep{kabsch1983dictionary} and normalized to a relative solvent accessibility using the absolute accessibilities from \citet{tien2013maximum}.
{\bf (C)} 
Sites of significant shifts are clustered in Env's structure.
Shown is the distance of each significantly shifted and not-shifted site to the closest other shifted site.
{\bf (D)}
Large mutational shifts are \emph{not} strongly enriched at sites that have substituted between BG505 and BF520, or at sites that contact sites that have substituted. 
The plot shows the magnitudes of the shifts among the 83 crystallographically resolved sites that have substituted between BG505 and BF520, the 234 non-substituted sites that physically contact a substitution in the Env structure (any non-hydrogen atom within 3.5\angstrom), and all other sites.
\FIGSUPP[shiftfeatures]{subs_shifts} shows that there is a modest borderline-significant tendency of significantly shifted sites to have substituted.
All panels only show the 577 sites that are resolved in the crystal structure.
Of the 30 significantly shifted sites in \FIG{prefsdist}B, only 28 are crystallographically resolved.
Structural distances and solvent accessibilities were calculated using all monomers in the trimer.
$P$ values were calculated using the Mann-Whitney $U$ test.
}
\figsupp[\label{figsupp:subs_shifts}Statistical testing of whether significantly shifted sites are more likely to have substituted between BG505 and BF520.]
{The sites of significant shifts as determined in \FIG{prefsdist}B are somewhat more likely to be among the sites that have substituted between BG505 and BF520.
However, this association is only borderline statistically significant, with $P = 0.055$ using a Fisher's exact test.}
{\input{figures/shifts_vs_subs_table}}
\end{figure}

What distinguishes the sites that have undergone significant shifts?
\FIG{shiftfeatures}A shows the locations of the shifted sites on the crystal structure of Env.
There is no visually obvious tendency for shifted sites to preferentially be on Env's surface or in its core, and a statistical analysis (\FIG{shiftfeatures}B) finds no association between a site's relative solvent accessibility and whether its amino-acid preferences have shifted.
However, \FIG{shiftfeatures}A does suggest that the sites of significant shifts tend to cluster in Env's structure, and a statistical analysis confirms that this is the case (\FIG{shiftfeatures}C).
Therefore, the factors that drive shifts in Env's mutational tolerance often affect physically interacting clusters of residues in a coordinated fashion.

An obvious hypothesis is that strongly shifted sites have substituted between BG505 and BF520, or physically contact such substitutions.
According to this hypothesis, substitutions would alter the local physicochemical environment of the substituted site and its neighbors, thereby shifting the amino-acid preferences of sites in the physical cluster.
But surprisingly, the typical magnitude of shifts is not significantly larger at sites that have substituted, or at sites that contact sites that have experienced substitutions (\FIG{shiftfeatures}C).
There is a borderline trend for the significantly shifted sites to be more likely to have substituted between BG505 and BF520 (\FIGSUPP[shiftfeatures]{subs_shifts}), but most shifted sites have not substituted (only 8 of the 30 shifted sites differ in amino-acid identity between the two Envs).
The lack of strong enrichment in shifts at substituted sites contrasts with previous protein-wide experimental~\citep{doud2015site} and simulation-based~\citep{pollock2012amino,shah2015contingency} studies of shifting amino-acid preferences, which found that shifts were dramatically more pronounced at sites of substitutions.
The difference may arise because these earlier studies examined proteins that are fairly conformationally static (absolutely so in the case of the simulations).
On the other hand, Env is extremely complex and conformationally dynamic~\citep{munro2014conformational,ozorowski2017open}, which may increase the opportunities for long-range epistasis to enable substitutions at one site to shift the amino-acid preferences of distant sites.

@@

\FIG{entrenchment} shows the distribution of effects for mutating all sites that differ between BG505 and BF520 to the identity in the other Env.
As expected under entrenchment, the average effect of these mutations is deleterious -- although there are a substantial number of sites where the mutational flips are not deleterious. 
We can get some sense of the magnitude of the entrenchment by comparing the effects of the BG505$\leftrightarrow$BF520 mutations to the distribution of effects of all possible amino-acid mutations (\FIG{entrenchment}).
This comparison shows that even unfavorable inter-Env mutational flips are generally more favorable than random amino-acid mutations.
Therefore, entrenchment occurs for some but not all substitutions that distinguish BG505 and BF520, and the magnitude of entrenchment is less than the effect of a typical random mutation.
Entrenchment of substitutions therefore contributes to some of the mutational shifts.
But given that many of these shifts occur at sites that do not even differ between the Envs (\FIG{shiftfeatures}D), entrenchment of substitutions is clearly not the only cause of the shifting amino-acid preferences.

@@

Our results show that amino-acid preference maps of Env also have a useful level of conservation for many purposes.
From a qualitative perspective, the amino-acid preferences look mostly similar between BG505 and BF520, and so provide a valuable reference for estimating which mutations are likely to be tolerated at each site in diverse HIV strains.
Indeed, we anticipate that the complete maps of mutational effects in \FIG{BG505prefs} and \FIG{BF520prefs} will be useful for future sequence-structure-function studies.
From an analytical perspective, a powerful use of our maps is to identify sites that evolve differently in nature than is required by the simple selection for viral growth imposed in our experiments -- and the identified sites are largely the same regardless of whether the analysis uses an amino-acid preference map from BG505 or BF520.
